---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  warning = FALSE, message = FALSE
)
```

# Overview of the package
This package allows you to select matched controls for an exposure variable under study in environmental epidemiology. Then you will be able to conduct a matched analysis to examine the health effects of the exposure variable. 

## Installation

You can install smurf from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
# devtools::install_github("my1120/smurf")
```

## Example

```{r packages}
library(easypackages)
libraries("smurf", "dlnm", "eesim", "dplyr", "data.table")
```

A single city dataset is available from the package "dlnm".
```{r}
data("chicagoNMMAPS")
chic <- chicagoNMMAPS
head(chic)
```

Simulate a multi-city dataset with continuous exposure variable.
```{r data simulation}
city_a <- create_sims(n_reps = 1, n = 365 * 5, exposure_type = "continuous",
                      exposure_trend = "cos1", central = 15, sd = 1,
                      exposure_amp = 0.6,
                      average_outcome = 40, outcome_trend = "cos1linear",
                      outcome_amp = 0.6,
                      rr = 1.2, start.date = "2000-01-01")
city_a <- do.call(rbind.data.frame, city_a)
city_a$city <- "A"

city_b <- create_sims(n_reps = 1, n = 365 * 5, exposure_type = "continuous",
                      exposure_trend = "cos1", central = 30, sd = 1,
                      exposure_amp = 0.6,
                      average_outcome = 40, outcome_trend = "cos1linear",
                      outcome_amp = 0.6,
                      rr = 1.4, start.date = "2000-01-01")
city_b <- do.call(rbind.data.frame, city_b)
city_b$city <- "B"

city_c <- create_sims(n_reps = 1, n = 365 * 5, exposure_type = "continuous",
                      exposure_trend = "cos1", central = 10, sd = 1,
                      exposure_amp = 0.6,
                      average_outcome = 40, outcome_trend = "cos1linear",
                      outcome_amp = 0.6,
                      rr = 1.02, start.date = "2000-01-01")
city_c <- do.call(rbind.data.frame, city_c)
city_c$city <- "C"

mcity <- rbind(city_a, city_b, city_c) %>%
  rename(exposure = x)
head(mcity)
```

#### findevent
Under some circumstances, an environmental exposure event is defined based on continuous measurements. For example, heatwave is typically defined as .... [citation]. "findevent" function can be used to find exposure event based on some conditions.

First, here is an example of finding exposure event for a single city. 
```{r}
chic_xmin <- data.table::data.table(byid = 1, xmin = 30)
chic_event <- findevents(date = chic$date, x = chic$temp, xmint = chic_xmin, 
                         mindays = 2)
subset(chic_event, event == TRUE)
```

Second, if the datasets contains data for multiple cities, you would like to find events within each city. Here is an example using the simulated multicities dataset.
```{r}
# find the tresholds within each city
m_xmin <- mcity %>%
  group_by(city) %>%
  summarise(xmin = quantile(exposure, probs = 0.8)) %>%
  mutate(byid = 1:length(city)) %>%
  select(byid, xmin) %>%
  as.data.frame()

mevent_1 <- findevents(date = mcity$date, x = mcity$exposure, 
                      xmint = m_xmin, 
                      mindays = 2, by = mcity$city)
mevent_1
```





